Astrophysical calibration of gravitational-wave detectors

### 2015-10-07 docs/cosmic_cal.tex
Added commnets from meeting.

\begin{abstract}
  We present an investigation into the potential of assessing the validity of the 
calibration of gravitational wave detector outputs through the use of standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  obtained and combined with current knowledge of the cosmological
  parameters.  This will yield the true luminosity distance and allow
  the direct comparison with the estimated value.  Discrepancies can
  then be used to correct the original calibration.~\cm{Update at end}
\end{abstract}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW} detectors will detect waves 
emitted from $O(10s)$ of \acp{CBC}. One such class of these cataclysmic events, the inspiral and
merger of \ac{BNS} systems will be detected out to a maximum range of $\approx 450$ Mpc. Assuming 
the current best estimates for the cosmological parameters, this is equivalent to a redshift 
$z\approx 0.1$. As noted by \cite{1986Natur.323..310S} \ac{CBC} systems can be used as cosmological 
distance markers, otherwise known as ``standard sirens'' (analogous to the \ac{EM} standard 
candles). Standard sirens allow ...

% BNS systems are GRBs
It is considered likely that the merger of \ac{BNS} and/or \ac{NSBH} systems, in addition to 
emitting detectable \acp{GW}, is also the mechanism for producing \acp{sGRB} 
\cite{1992ApJ...395L..83N}. In this scenario these events produce tightly beamed \ac{EM} emission 
parallel to the orbital angular 
momentum vector of the \ac{BNS} system. Additional evidence for the coincidence of \acp{sGRB} with
\ac{GW} events are the estimated astrophysical rates of both phenomena. Observations of \acp{sGRB} 
give a rate of XXX\,Mpc$^{-3}$\,yr$^{-1}$ which can be compared to estimates of \ac{BNS} merger 
rates of $10^{-8}$--$10^{-5}$\,Mpc$^{-3}$\,yr$^{-1}$ \cite{2010CQGra..27q3001A}, or \ac{NSBH} merger 
rates of $6\,\times\,10^{-10}$--$1\,\times\,10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$ 
\cite{2010CQGra..27q3001A}. However, the recent work of \cite{2015MNRAS.448..928K}, which studied 
known galactic \ac{BNS} systems in more detail suggest a rate approximately five times lower. The 
estimates for \ac{NSBH} systems come entirely from population synthesis models as no such systems 
have been observed.
%The ranges of rates for \acp{BNS} is obtained from population synthesis models together with 
%knowledge of the X known \ac{BNS} systems in our galaxy, whilst the range for \acp{NSBH} is 
%entirely from population synthesis as not such systems have been observed.

@@

For future planned space-based detectors such as eLISA there are several known galactic binary 
systems known as ``verification'' binaries \cite{2006CQGra..23S.809S}, in that their masses and 
orbital parameters allow them to be guarenteed sources for such detectors. These too could be used 
for assessing calibiration for such detectors. However, generally for these verification binaries 
the inclination of the system is not known and distances are fairly uncertain, so an accurate 
calibration assessment may not be possible. Although there is at least one (J0651) currently known 
eclipsing white dwarf binary \cite{2041-8205-737-1-L23}, which does provide the system inclination 
\cite{2012ApJ...757L..21H, 2013ASPC..467...47K} and has an approximately 10\% distance uncertainty 
\cite{2041-8205-737-1-L23}, which make it an excellent candiate for an astrophysical calibrator at 
its frequency.

@@

% continued use of sirens
The use of \ac{GW} events with \ac{sGRB} counterparts for use as a cosmological tool was 
investigated in~\cite{2010CQGra..27u5006S} and~\cite{2011PhRvD..83b3005Z} with respect to
the 3rd generation \ac{GW} interferometer, the \ac{ET}. It was
assumed in both studies that 1000 coincident events would be observed and
cosmological parameter estimation ability was compared with current \ac{EM}
results with focus on dark energy parameters. Other methods for cosmological
inference (see~\cite{1996PhRvD..53.2878F,2012PhRvD..85b3535T, 2012PhRvD..86b3502T}) have 
been proposed for \ac{BNS} systems that do
not rely on any redshift measurements and instead use the distribution of
measured \acp{SNR} combined with some assumptions regarding the \ac{NS} mass
and spatial distributions. Finally, we mention that for \ac{BNS} systems it has
been shown in~\cite{Messenger:2011ux} and~\cite{2013arXiv1312.1862M} that features
of the tidal and post-merger hyper-massive \ac{NS} stages of the \ac{GW}
waveform can allow the measurement of the source redshift if the \ac{NS}
equation-of-state is sufficiently well known. This development, and separately, the method
based on \ac{SNR} distributions will allow \ac{GW} cosmological inference
without the need for complementary \ac{EM} observations.

@@

% discuss joint-event rates
The predicted detection rates of \ac{BNS} events (irrespective of an \ac{sGRB}
counterpart) are derived from three main sources, population synthesis models,
extrapolations based on known galactic \ac{BNS} systems, and from observations
of \ac{sGRB} events. Various studies using the first two methods have been
compiled into an overall detection rate of $0.2$--$200$ \ac{BNS} events per
year~\cite{2013arXiv1304.0670L,2010CQGra..27q3001A}\footnote{We note that this rate is that given 
in Table 1 of \cite{2013arXiv1304.0670L} for an \ac{aLIGO} and \ac{AdV} network consisting to two 
\ac{aLIGO} detectors operating at design sensitivity. However, a higher rate of $0.4$--$400$ is 
given in \cite{2010CQGra..27q3001A}, which may differ due to the use of a third \ac{aLIGO} 
detector, but the discrepancy may also have contributions from slightly different threshold 
definitions or design sensitivity curves used. We also note the study mentioned in 
Sec.~\ref{sec:intro} that potentially lowers the expected rate by a factor of five.}. 
In~\cite{2012MNRAS.425.2668C} using the rate of observed \acp{sGRB} and
assuming a beaming angle of $15^{o}$ a similar event rate of $1$--$180$ is
obtained. The beaming angle is the primary factor in determining the fraction
of detected \ac{GW} \ac{BNS} signals with an \ac{EM} \ac{sGRB} counterpart and
is unfortunately highly uncertain.  Various authors have provided estimates on
the rate of joint detections to be in the range $0.02$--$7$ per
year~\cite{2013PhRvL.111r1101C,Kelley:2012fl,2014arXiv1405.2254S,2015MNRAS.448.3026W}.

@@

% discuss the inclination and opening angle
Other information relevant for a coincident \ac{sGRB} is that the system is
likely to be relatively close to face-on (and therefore nearly circularly
polarised), which allows us to place prior constraints on the system
inclination angle. Beam opening angles for \acp{sGRB} are relatively poorly
constrained as they are based on only a few jet-break detections. A comparative
study of estimates and lower-limits of opening angles in 13 \acp{sGRB} given
in~\cite{2014ApJ...780..118F} provide a median opening angle value of $\sim
10^{\circ}$. We assume that the beam opening angle is a half-normal
distribution, for which the median is given by $\sqrt{2}\erf{}^{-1}(1/2)\sigma
\sim 0.67\sigma$. It follows that the median of a $\sim 10^{\circ}$ opening
angle corresponds to $\sigma \sim 14.8^{\circ}$. We therefore place a Gaussian
prior on the inclination angle $\iota$ with zero mean and a standard deviation
$\sigma=14.8^{\circ}$.

@@

% the scale factor prior
Finally, we require a prior on the calibration scale factors for each detector.
We make the assumption that in general the calibration applied to the detector
data will be correct, so want to use a prior that is peaked at $\scf_k=1$. We
also assume that the prior \ac{PDF} of the calibration scale factor is either
larger or smaller than unity by an equivalent factor, e.g.\ $p(\scf|I)=10$ is
equivalent to $p(\scf|I)=0.1$. For the scale factor in each detector we 
therefore use the following log-normal distribution as our prior
%
\begin{equation}\label{eq:scale_prior}
 p(\scf|I) = \frac{1}{\scf\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{\scf} - \mu)^2}{2\sigma^2} 
\right)},
\end{equation}
%
where we chose $\sigma = 1.07$, which together with the requirement that the
mode is at unity gives $\mu = 1.15$. With these parameters the prior
probability density at 0.1 and 10 is one tenth of that at 1. Note that this
prior does not give a symmetric amount of probability about unity (so out prior favour \scf\xspace 
values greater than 1), but given
our simulation criterion described in Section~\ref{sec:simulations} we find
that the likelihood generally overwhelms the prior~\cm{I'm not sure about this
sentence.  Why does the symmetry matter?}. As $\scf$ is a scale factor we also
have the further constraint that it must be positive and hence the log-normal
prior is a natural choice.

@@

% a discussion of the PSD estimate
For each of the simulated sources we have used the {\tt emcee} python \ac{MCMC}
package~\cite{2013PASP..125..306F} to perform parameter estimation over the
unknown source parameters using the priors as discussed in
Section~\ref{sec:priors}. In each case when calculating the likelihood we used
an estimate of the noise \ac{PSD} based on the advanced detector design
sensitivities (using those given in~\cite{2013arXiv1304.0670L}), but calculated by averaging 
64 separate noisy realisations of the \ac{PSD} and scaled
with the same factor as applied to the injection and noise\footnote{We do not account for there 
being a potential difference
between the estimated \ac{PSD} and the actual \ac{PSD} of the analysed section
of data as described in e.g.\ \cite{2013PhRvD..88h4044L}. This difference would
be very highly correlated with the calibration scale factor, so in reality our
estimate of the scale factor would be a combination of the calibration offset
and the difference in the \ac{PSD}. As such our results on real data would be
an upper limit on the calibration scale factor.~\cm{Even though we're hiding
this in a footnote is it really this bad. How big is the statistical error on
PSD estimation using 64 segments?}}. This has provided posterior probability
distributions on the calibration scale factors for each detector.  Examples of
the marginalised posterior probabilities for a \ac{BNS} system and a \ac{NSBH}
system observed with the three detector network are shown in
figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

@@

% discuss the NSBH sources
In figure~\ref{fig:nsbhresults} we see very similar results for the \ac{NSBH}
systems although the higher \ac{SNR} of the signals means that we can provide
$\sim 20\mbox{--}25\%$ uncertainties on the calibration scale factors for H1
and L1 out to greater distances. The distributions generally appear slightly
broader than those for the \ac{BNS} systems. This may be explained
by two factors. The first is from the fact that including the spin
parameter leads to strong correlations between the chirp mass, mass ratio and
spin. These strong correlations make convergence of the \ac{MCMC} take longer
and means we have fewer independent samples with which to estimate the posterior
distributions. This leads to a larger statistical fluctuation on the
results.  The second factor is that there appears to be a modest effect due to
the population of sources that are detectable. For distances at which the
\ac{BNS} and \ac{NSBH} systems would give the same percentage of observable
sources there is a slight increase in the mean and standard deviation of
\acp{SNR} of the \ac{NSBH} systems over the \ac{BNS} systems~\cm{can we go any
further with this?}.

@@

% mention caveats and future work
The requirement of a coincident \ac{sGRB} with a known distance means that
there will likely be considerable delay (beyond initial \ac{GW} detection) for
the implementation of this method of calibration assessment.  Importantly
however, this method would provide a semi-independent consistency check that
the calibration is reasonably accurate.  We acknowledge that we have made a
number of simplifying assumptions, including that the waveforms we use
accurately model the true signal, and that the overall calibration remains
constant over the duration of a signal and its \ac{PSD} estimation. A further
major simplification made here is that the absolute calibration is constant
with frequency.  Other authors (e.g.\ \cite{Vitale:2012,2013PhRvD..88h4044L} have addressed
this issue in the context of robust parameter estimation for astrophysical
sources and in future work we intend to apply related methods for astrophysical
calibration.

@@

\MP{Chris: feel free to add a short paragraph on relative calibration here.}

>>>>

\begin{abstract}
  We present an investigation assessing the validity of the 
of gravitational wave detector calibration through the use of gamma-ray
bursts as standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  combined with current knowledge of the cosmological
  parameters yielding the true luminosity distance.  This will then allow
  a direct comparison with the estimated value and discrepancies can
  then be used to correct the original calibration.~\MP{Update with results.
End with a statement comparing to existing method but stressing its independent.}
\end{abstract}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW} detectors will detect waves 
emitted from $O(10s)$ of \acp{CBC}. One such class of these cataclysmic events, the inspiral and
merger of \ac{BNS} systems will be detected out to a maximum range of $\approx 450$ Mpc. Assuming 
the current best estimates for the cosmological parameters, this is equivalent to a redshift 
$z\approx 0.1$. As noted by \cite{1986Natur.323..310S} \ac{CBC} systems can be used as cosmological 
distance markers, otherwise known as ``standard sirens'' (analogous to the \ac{EM} standard 
candles). \cm{Standard sirens allow ... fix me}

% BNS systems are GRBs
It is considered likely that the merger of \ac{BNS} and/or \ac{NSBH} systems, in addition to 
emitting detectable \acp{GW}, is also the mechanism for producing \acp{sGRB} 
\cite{1992ApJ...395L..83N}. In this scenario these events produce tightly beamed \ac{EM} emission 
parallel to the orbital angular 
momentum vector of the \ac{BNS} system. Additional evidence for the coincidence of \acp{sGRB} with
\ac{GW} events are the estimated astrophysical rates of both phenomena. Observations of \acp{sGRB} 
give a rate of \cm{XXX}\,Mpc$^{-3}$\,yr$^{-1}$ which can be compared to estimates of \ac{BNS} merger 
rates of $10^{-8}$--$10^{-5}$\,Mpc$^{-3}$\,yr$^{-1}$ \cite{2010CQGra..27q3001A}, or \ac{NSBH} merger 
rates of $6\,\times\,10^{-10}$--$1\,\times\,10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$ 
\cite{2010CQGra..27q3001A}. However, the recent work of \cite{2015MNRAS.448..928K}, which studied 
known galactic \ac{BNS} systems in more detail suggest a rate approximately five times lower. The 
estimates for \ac{NSBH} systems come entirely from population synthesis models as no such systems 
have yet been observed.
%The ranges of rates for \acp{BNS} is obtained from population synthesis models together with 
%knowledge of the X known \ac{BNS} systems in our galaxy, whilst the range for \acp{NSBH} is 
%entirely from population synthesis as not such systems have been observed.

@@

\MP{Move this paragraph to conclusions} For future planned space-based detectors such as eLISA there are several known galactic binary 
systems known as ``verification'' binaries \cite{2006CQGra..23S.809S}, in that their masses and 
orbital parameters allow them to be guarenteed sources for such detectors. These too could be used 
for assessing calibiration for such detectors. However, generally for these verification binaries 
the inclination of the system is not known and distances are fairly uncertain, so an accurate 
calibration assessment may not be possible. Although there is at least one (J0651) currently known 
eclipsing white dwarf binary \cite{2041-8205-737-1-L23}, which does provide the system inclination 
\cite{2012ApJ...757L..21H, 2013ASPC..467...47K} and has an approximately 10\% distance uncertainty 
\cite{2041-8205-737-1-L23}, which make it an excellent candiate for an astrophysical calibrator at 
its frequency.

@@

% continued use of sirens
\cm{Lose, trim or merge this paragraph with previous one.} The use of \ac{GW} events with \ac{sGRB} counterparts for use as a cosmological tool was 
investigated in~\cite{2010CQGra..27u5006S} and~\cite{2011PhRvD..83b3005Z} with respect to
the 3rd generation \ac{GW} interferometer, the \ac{ET}. It was
assumed in both studies that 1000 coincident events would be observed and
cosmological parameter estimation ability was compared with current \ac{EM}
results with focus on dark energy parameters. Other methods for cosmological
inference (see~\cite{1996PhRvD..53.2878F,2012PhRvD..85b3535T, 2012PhRvD..86b3502T}) have 
been proposed for \ac{BNS} systems that do
not rely on any redshift measurements and instead use the distribution of
measured \acp{SNR} combined with some assumptions regarding the \ac{NS} mass
and spatial distributions. Finally, we mention that for \ac{BNS} systems it has
been shown in~\cite{Messenger:2011ux} and~\cite{2013arXiv1312.1862M} that features
of the tidal and post-merger hyper-massive \ac{NS} stages of the \ac{GW}
waveform can allow the measurement of the source redshift if the \ac{NS}
equation-of-state is sufficiently well known. This development, and separately, the method
based on \ac{SNR} distributions will allow \ac{GW} cosmological inference
without the need for complementary \ac{EM} observations.

@@

% discuss joint-event rates
The predicted detection rates of \ac{BNS} events (irrespective of an \ac{sGRB}
counterpart) are derived from three main sources, population synthesis models,
extrapolations based on known galactic \ac{BNS} systems, and from observations
of \ac{sGRB} events. Various studies using the first two methods have been
compiled into an overall detection rate of $0.2$--$200$ \ac{BNS} events per
year~\cite{2013arXiv1304.0670L,2010CQGra..27q3001A}\footnote{\cm{Email Patrick
S about this} We note that this rate is that given 
in Table 1 of \cite{2013arXiv1304.0670L} for an \ac{aLIGO} and \ac{AdV} network consisting to two 
\ac{aLIGO} detectors operating at design sensitivity. However, a higher rate of $0.4$--$400$ is 
given in \cite{2010CQGra..27q3001A}, which may differ due to the use of a third \ac{aLIGO} 
detector, but the discrepancy may also have contributions from slightly different threshold 
definitions or design sensitivity curves used. We also note the study mentioned in 
Sec.~\ref{sec:intro} that potentially lowers the expected rate by a factor of five.}. 
In~\cite{2012MNRAS.425.2668C} using the rate of observed \acp{sGRB} and
assuming a beaming angle of $15^{o}$ a similar event rate of $1$--$180$ is
obtained. The beaming angle is the primary factor in determining the fraction
of detected \ac{GW} \ac{BNS} signals with an \ac{EM} \ac{sGRB} counterpart and
is unfortunately highly uncertain.  Various authors have provided estimates on
the rate of joint detections to be in the range $0.02$--$7$ per
year~\cite{2013PhRvL.111r1101C,Kelley:2012fl,2014arXiv1405.2254S,2015MNRAS.448.3026W}.

@@

% discuss the inclination and opening angle
Other information relevant for a coincident \ac{sGRB} is that the system is
likely to be relatively close to face-on (and therefore nearly circularly
polarised), which allows us to place prior constraints on the system
inclination angle. Beam opening angles for \acp{sGRB} are relatively poorly
constrained as they are based on only a few jet-break detections. A comparative
study of estimates and lower-limits of opening angles in 13 \acp{sGRB} given
in~\cite{2014ApJ...780..118F} provide a median opening angle value of $\sim
10^{\circ}$. We assume that the beam opening angle is a half-normal
distribution, for which the median is given by $\sqrt{2}\erf{}^{-1}(1/2)\sigma
\sim 0.67\sigma$. It follows that the median of a $\sim 10^{\circ}$ opening
angle corresponds to $\sigma \sim 14.8^{\circ}$. We therefore place a Gaussian
prior on the inclination angle $\iota$ with zero mean and a standard deviation
$\sigma=14.8^{\circ}$.~\MP{Check regarding the face/arse orientation issue.}

@@

% the scale factor prior
Finally, we require a prior on the calibration scale factors for each detector.
We make the assumption that in general the calibration applied to the detector
data will be correct, so want to use a prior that is peaked at $\scf_k=1$. We
also assume that the prior \ac{PDF} of the calibration scale factor is either
larger or smaller than unity by an equivalent factor, e.g.\ $p(\scf|I)=10$ is
equivalent to $p(\scf|I)=0.1$. For the scale factor in each detector we 
therefore use the following log-normal distribution as our prior
%
\begin{equation}\label{eq:scale_prior}
 p(\scf|I) = \frac{1}{\scf\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{\scf} - \mu)^2}{2\sigma^2} 
\right)},
\end{equation}
%
where we chose $\sigma = 1.07$, which together with the requirement that the
mode is at unity gives $\mu = 1.15$. With these parameters the prior
probability density at 0.1 and 10 is one tenth of that at 1. Note that this
prior does not give a symmetric amount of probability about unity (so our prior
favour \scf\xspace values greater than 1), but given our simulation criterion
described in Section~\ref{sec:simulations} we find that the likelihood
generally overwhelms the prior. As $\scf$ is a scale factor we also have the
further constraint that it must be positive and hence the log-normal prior is a
natural choice.

@@

% a discussion of the PSD estimate
For each of the simulated sources we have used the {\tt emcee} python \ac{MCMC}
package~\cite{2013PASP..125..306F} to perform parameter estimation over the
unknown source parameters using the priors as discussed in
Section~\ref{sec:priors}. In each case when calculating the likelihood we used
an estimate of the noise \ac{PSD} based on the advanced detector design
sensitivities (using those given in~\cite{2013arXiv1304.0670L}), but calculated by averaging 
64 separate noisy realisations of the \ac{PSD} and scaled
with the same factor as applied to the injection and noise\footnote{We do not account for there 
being a potential difference
between the estimated \ac{PSD} and the actual \ac{PSD} of the analysed section
of data as described in e.g.\ \cite{2013PhRvD..88h4044L}. This difference would
be very highly correlated with the calibration scale factor, so in reality our
estimate of the scale factor would be a combination of the calibration offset
and the difference in the \ac{PSD}. As such our results on real data would be
an upper limit on the calibration scale factor.}. This has provided posterior probability
distributions on the calibration scale factors for each detector.  Examples of
the marginalised posterior probabilities for a \ac{BNS} system and a \ac{NSBH}
system observed with the three detector network are shown in
figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

@@

% discuss the NSBH sources
In figure~\ref{fig:nsbhresults} we see very similar results for the \ac{NSBH}
systems although the higher \ac{SNR} of the signals means that we can provide
$\sim 20\mbox{--}25\%$ uncertainties on the calibration scale factors for H1
and L1 out to greater distances. The distributions generally appear slightly
broader than those for the \ac{BNS} systems. This may be explained
by two factors. The first is from the fact that including the spin
parameter leads to strong correlations between the chirp mass, mass ratio and
spin. These strong correlations make convergence of the \ac{MCMC} take longer
and means we have fewer independent samples with which to estimate the posterior
distributions. This leads to a larger statistical fluctuation on the
results.  The second factor is that there appears to be a modest effect due to
the population of sources that are detectable. For distances at which the
\ac{BNS} and \ac{NSBH} systems would give the same percentage of observable
sources there is a slight increase in the mean and standard deviation of
\acp{SNR} of the \ac{NSBH} systems over the \ac{BNS} systems.

@@

% mention caveats and future work
The requirement of a coincident \ac{sGRB} with a known distance means that
there will likely be considerable delay (beyond initial \ac{GW} detection) for
the implementation of this method of calibration assessment.  Importantly
however, this method would provide a \MP{semi-independent} consistency check that
the calibration is reasonably accurate.  We acknowledge that we have made a
number of simplifying assumptions, including that the waveforms we use
accurately model the true signal, and that the overall calibration remains
constant over the duration of a signal and its \ac{PSD} estimation. A further
major simplification made here is that the absolute calibration is constant
with frequency.  Other authors (e.g.\ \cite{Vitale:2012,2013PhRvD..88h4044L} have addressed
this issue in the context of robust parameter estimation for astrophysical
sources and in future work we intend to apply related methods for astrophysical
calibration.

@@

\cm{Chris: feel free to add a short paragraph on relative calibration here.}
